---
title: "Shared Biological Pathways"
author: "Siwei Zhang"
date: "2024-07-09"
output: html_document
---

## Shared pathways module

```{r,echo=F,message=F,warning=F}
library(r2d3)
library(tidyverse)

dat = readRDS("enrichment_network_initial_update.rds")



```

## Shared pathways identified from common disease comorbidities (ICD-based) based on public resource

- Searching by disease (ICD10 diagnosis code) and identify the associated genes based on the disgenet2r database.

- Pathway enrichment analysis on the associated genes

```{r,echo=F,message=F,warning=F}
dat = readRDS("disease_pairs_pathways_disgenet2r.rds") %>%
  filter(pathways_common!=" ") %>%
  dplyr::rename(shared_pathways = pathways_common)
DT::datatable(dat)

```

## Shared pathways identified from common disease comorbidities (ICD-based) based on Phe-OmicsMIME

- Identified from ICD-based disease diagnosis phenotype to shared pathways.

```{r,echo=F,message=F,warning=F}
# library(phewasHelper)
# disease_pairs <- tibble(
#   disease1 = c(
#     "Cardiovascular Disease", "Obesity", "Alzheimer's Disease", "Chronic Obstructive Pulmonary Disease",
#     "Malignant Neoplasms", "Rheumatoid Arthritis", "Chronic Kidney Disease", "Essential hypertension",
#     "Asthma", "Depression", "Parkinson's Disease", "Psoriasis", "Migraine", "Multiple Sclerosis",
#     "Type 1 Diabetes", "Hyperlipidemia"
#   ),
#   phecode1 = c(
#     "401", "278.1", "290.1", "496",
#     "174", "714", "585", "401.1",
#     "493", "296.2", "332", "696.1", "346", "335",
#     "250.1", "272.1"
#   ),
#   phenotype1 = c(
#     "Essential hypertension", "Obesity", "Alzheimer's disease", "Chronic airway obstruction",
#     "Malignant neoplasms", "Rheumatoid arthritis", "Chronic renal failure [CKD]", "Essential hypertension",
#     "Asthma", "Major depressive disorder", "Parkinson's disease", "Psoriasis", "Migraine", "Multiple sclerosis",
#     "Type 1 diabetes", "Hyperlipidemia"
#   ),
#   icd10_code1 = paste0("ICD10_",c(
#     "I10", "E66", "G30", "J44",
#     "C00-C97", "M06.9", "N18", "I10",
#     "J45", "F32", "G20", "L40", "G43", "G35",
#     "E10", "E78"
#   )),
#   doid1 = c(
#     "DOID:10763", "DOID:9970", "DOID:10652", "DOID:3083",
#     "DOID:162", "DOID:7148", "DOID:1220", "DOID:10763",
#     "DOID:2841", "DOID:150", "DOID:14330", "DOID:8893", "DOID:6364", "DOID:2377",
#     "DOID:9744", "DOID:2355"
#   ),
#   h_number1 = c(
#     "H01633","H02106","H00056","H01714"," ","H00630"," ","H01633","H00079","H01646","H00057","H01656","H00775","H01490","H00408","H01635"
#   ),
#   disease2 = c(
#     "Type 2 diabetes", "Type 2 Diabetes", "Cardiovascular Disease", "Cardiovascular Disease",
#     "Obesity", "Cardiovascular Disease", "Cardiovascular Disease", "Stroke",
#     "Allergic Rhinitis", "Anxiety Disorders", "Alzheimer's Disease", "Rheumatoid Arthritis", "Fibromyalgia", "Systemic Lupus Erythematosus",
#     "Celiac Disease", "Coronary Artery Disease"
#   ),
#   phecode2 = c(
#     "250", "250.2", "401", "401",
#     "278.1", "401", "401", "433",
#     "477", "300", "290.1", "714", "729.1", "710.2",
#     "579.0", "411"
#   ),
#   phenotype2 = c(
#     "Type 2 diabetes", "Type 2 diabetes", "Essential hypertension", "Essential hypertension",
#     "Obesity", "Essential hypertension", "Essential hypertension", "Occlusion of cerebral arteries",
#     "Allergic rhinitis", "Anxiety disorder", "Alzheimer's disease", "Rheumatoid arthritis", "Fibromyalgia and myofascial pain syndromes", "Systemic lupus erythematosus",
#     "Celiac disease", "Coronary atherosclerosis"
#   ),
#   icd10_code2 = paste0("ICD10_",c(
#     "E11", "E11", "I10", "I10",
#     "E66", "I10", "I10", "I63",
#     "J30", "F41", "G30", "M06.9", "M79.7", "M32",
#     "K90.0", "I25"
#   )),
#   doid2 = c(
#     "DOID:9352", "DOID:9352", "DOID:10763", "DOID:10763",
#     "DOID:9970", "DOID:10763", "DOID:10763", "DOID:1508",
#     "DOID:0060509", "DOID:2841", "DOID:10652", "DOID:7148", "DOID:2657", "DOID:6757",
#     "DOID:10608", "DOID:3393"
#   ),
#   h_number2 = c("H00409","H00409","H01633","H01633","H02106","H01633","H01633"," ","H01360","H01662","H00056","H00630","H01636","H00080","H02123","H02505")
# ) %>% mutate(phecode1 = normalize_phecodes(phecode1),phecode2 = normalize_phecodes(phecode2))
# 
# # current_institution = "vumc"
# res = purrr::map2_dfr(disease_pairs$phenotype1, disease_pairs$phenotype2, function(x,y){
# 
#   conn_nodes = tidy_connect %>%
#           # filter(institution == current_institution) %>%
#           filter(from %in% x) %>%
#           dplyr::select(node = to,hazard_ratio) %>%
#           bind_rows(tidy_connect %>%
#                       filter(institution == current_institution) %>%
#                       filter(to %in% x) %>%
#                       dplyr::select(node = from,hazard_ratio)) %>%
#           distinct(node,.keep_all = T) %>%
#           mutate(phenotype = x) %>%
#           bind_rows(tidy_connect %>%
#           filter(institution == current_institution) %>%
#           filter(from %in% y) %>%
#           dplyr::select(node = to,hazard_ratio) %>%
#           bind_rows(tidy_connect %>%
#                       filter(institution == current_institution) %>%
#                       filter(to %in% y) %>%
#                       dplyr::select(node = from,hazard_ratio)) %>%
#           distinct(node,.keep_all = T) %>%
#           mutate(phenotype = y))
#   
#   if(nrow(conn_nodes)!=0){
#   
#     dat = conn_nodes %>%
#         left_join(.,all_id %>% dplyr::rename(node = variable) %>% dplyr::select(node,orig_id) %>%
#                     filter(!is.na(node)) %>% distinct(node,.keep_all = T), by="node") %>%
#         left_join(.,nodes %>% filter(institution==current_institution),by="node")
#   dat = convert_id(dat)
#   res = purrr::map_dfr(unique(dat$phenotype),function(pheno){
#      gene_list = dat %>% filter(phenotype==pheno) %>%
#        distinct(orig_id,.keep_all = T) %>%
#        distinct(ENTREZID,.keep_all = T) %>%
#        distinct(node,.keep_all = T) %>%
#        arrange(desc(hazard_ratio)) %>%
#        pull(hazard_ratio)
#      names(gene_list) = dat %>% filter(phenotype==pheno) %>%
#        distinct(orig_id,.keep_all = T) %>%
#        distinct(ENTREZID,.keep_all = T) %>%
#        distinct(node,.keep_all = T) %>%
#        pull(ENTREZID)
# 
#      ## go enrichment analysis
#      res = gseGO(geneList   = gene_list,
#                  OrgDb        = org.Hs.eg.db,
#                  ont          = "CC",
#                  minGSSize    = 10,
#                  # maxGSSize    = 500,
#                  pvalueCutoff = 1,
#                  verbose      = FALSE)
#      res_go = res@result
# 
#      ## kegg enrichment analysis
#      res = gseKEGG(geneList     = gene_list,
#                    organism     = 'hsa',
#                    minGSSize    = 10,
#                    pvalueCutoff = 1,
#                    verbose      = FALSE)
#      res_kegg = res@result
# 
#      ## wikipathway analysis
#      # res = gseWP(gene_list,
#      #             organism = "Homo sapiens",
#      #             minGSSize    = 100,
#      #             pvalueCutoff = 1,
#      #             verbose      = FALSE)
#      # res_wiki = res@result
#      #
#      ## reactome pathway
#      res <- gsePathway(gene_list,
#                      minGSSize    = 10,
#                      pvalueCutoff = 1,
#                      pAdjustMethod = "BH",
#                      verbose = FALSE)
#      res_reactome = res@result
# 
#      ## disease
#      res <- gseDO(gene_list,
#                 minGSSize     = 10,
#                 pvalueCutoff  = 1,
#                 pAdjustMethod = "BH",
#                 verbose       = FALSE)
#      res_disease = res@result
# 
#      bind_rows(res_go,res_kegg,res_reactome,res_disease) %>%
#        mutate(pvalue = round(pvalue,2),
#               p.adjust = round(p.adjust,2),
#               qvalue = round(qvalue,2),
#               enrichmentScore = round(enrichmentScore,2),
#               NES = round(NES,2)) %>%
#        mutate(phenotype = pheno)
#     })
# 
#   disease1 = res %>% filter(phenotype == x)
#   disease2 = res %>% filter(phenotype == y)
# 
#   if(nrow(disease1)!=0 & nrow(disease2)!=0){
#     pathways_common = intersect(disease1$Description,disease2$Description)
#   } else{
#     pathways_common = " "
#   }
# 
#   if(length(pathways_common)!=0){
#     pathways_common = data.frame(pathways_common) %>%
#       mutate(disease1 = x,
#              disease2 = y)
#   } else{
#     pathways_common = data.frame(pathways_common = " ",
#                                  disease1=x,
#                                  disease2=y)
#   }
#     
#   }
#   
#   pathways_common
# 
# })
# saveRDS(res, file="disease_pairs_pathways_pheomics.rds")
```

```{r,echo=F,message=F,warning=F}
dat2 = readRDS("disease_pairs_pathways_pheomics.rds") %>%
  filter(pathways_common!=" ") %>%
  dplyr::rename(shared_pathways = pathways_common)
DT::datatable(dat2)
# rsconnect::deploySite(siteName = "Phe-OmicsMIME_supplementary",account = "siwei",server="prod.tbilab.org")
```

## Targeted Disease Discovery Using Phe-OmicsMIME

### Type 2 diabetes and Hypertension

- VUMC cohort example

```{r,echo=F,message=F,warning=F}
pathway_diabetes_hypertension = readRDS("pathway_vumc_diabetes_hypertension.rds")

r2d3(
      data=pathway_diabetes_hypertension,
      script = "pathway_network.js",
      # dependencies = "inst/style.css",
      options = list(r2d3.theme = list(background="white")),
      container = "svg",
      d3_version = "5"
    )

DT::datatable(pathway_diabetes_hypertension$res_table)
```

- UKB cohort example

```{r,echo=F,message=F,warning=F}
pathway_diabetes_hypertension = readRDS("pathway_ukb_diabetes_hypertension.rds")

r2d3(
      data=pathway_diabetes_hypertension,
      script = "pathway_network.js",
      # dependencies = "inst/style.css",
      options = list(r2d3.theme = list(background="white")),
      container = "svg",
      d3_version = "5"
    )

DT::datatable(pathway_diabetes_hypertension$res_table)
```

### Alzheimer's disease and Parkinson's disease

- VUMC cohort example

```{r,echo=F,message=F,warning=F}
pathway_az_parkinson = readRDS("pathway_vumc_az_parkinson.rds")

r2d3(
      data=pathway_az_parkinson,
      script = "pathway_network.js",
      # dependencies = "inst/style.css",
      options = list(r2d3.theme = list(background="white")),
      container = "svg",
      d3_version = "5"
    )

DT::datatable(pathway_az_parkinson$res_table)
```


## Shared Biological Pathways in Consistent Disease Comorbidity Across Multiple Institutions Identified by PheMIME









